Gradual Soundness: Lessons from Static Python

### 2021-12-18 pj-7.1/paper.tex

The \SP{} solution is to provide two versions of every parameterized type,
both of which support efficient run-time checks.
One version is shallow.
The other is \emph{concrete}: complete, but new.
Consider \code{List[T]} as an example:
\begin{enumerate}
  \item
    \emph{Shallow type \code{List}}.
    The default type \code{List[T]} gets enforced with a simple check for lists that ignores
    the elements.
    Consequently, every static type like \code{List[int]} is merely a promise of \code{List}
    values at run-time.
  \item
    \emph{Concrete type \code{ChkList[T]}}.
    \SP{} introduces a new data structure and parameterized type \code{ChkList[T]}
    for concrete types.
    The data structure implements the list API and comes with a checked constructor that
    makes it impossible to create a \code{ChkList[int]} with non-integer elements;
    \cref{s:chkdict-impl} discusses the implementation further.
\end{enumerate}

@@

\subsubsection{Non-Highlights / Restrictions}

The \SP{} type system excludes include some features.
Each of the following is low priority for the team; that said, they are
ordered below rougly from most important (soonest to be implemented) to least.

@@

  \item \emph{Recursive Types}.
    \SP{} does not support recursive types.
    Neither does Pyre or mypy.
    %% demand seems low ... baffling
    %% mypy = https://github.com/python/mypy/issues/731
    %% nothing in pep 484 = https://www.python.org/dev/peps/pep-0484/
\end{itemize}

\subsection{Runtime System Highlights}
%% Non-Highlights:
%% - shadow frames = JIT feature, lite array to track Python call stack
%% - strict modules ... avoids toplevel function wrapper

\subsubsection{Bytecode Optimization}

optimizations and bytecodes for: length, subscripting, function dispatch, method dispatch, field load/store

table much later in paper

IN ADDITION to optimizations ...
not sure where this belongs exactly, but we go to a lot of work in the compiler
to intelligently minimize runtime checks by only inserting them when the source
value is of dynamic type (this includes also skipping CHECK\_ARGS if the
function was statically invoked by another static function and thus we know the
compiler checked argument types)

\subsubsection{Virtual Method Tables}

Cinder adds virtual method tables to typed classes.
``The v-tables are only used for the \code{INVOKE\_METHOD} opcode, which
the static compiler will only emit against a statically known static type.'' -Carl

with vtables, method resolution happens differently than Python: args. get
resolved before the receiver (A.m(B) goes "A->B" in Py. and "B->A" in SP)

vtables are built on Python's vectorcall signature; though new, they're not
totally from scratch

@@

\subsubsection{Checked Data Structures}
\label{s:chkdict-impl}

As mentioned above~(\cref{s:checked-type}), the concrete versions of built-in data structures
come with both a type and an implementation.
The implementation provides the same interface as the built-in;
the main difference is that it adds type checks.

For example, the type \code{ChkDict[K, V]} describes a concrete dictionary
with keys of type \code{K} and values of type \code{V}.
The implementation has three main components:
\begin{itemize}
  \item
    Constructor, basically from dict to CD, checks integrity.
    All CD valuess have to go through this, whether types or untyped.
  \item
    Instances keep a tag, used as exact type in boundary check.
    (Tag may not be CD[i i] but rather a sentinel generated from it.)
  \item
    Writes are protected in every context.
\end{itemize}

%    Each instance of a \code{ChkList[int]} comes with the full type as a tag;
%    run-time checks inspect this tag for an exact match.\footnote{Immutable
%    data structures could use a subtype check. \SP{} does not have any of these
%    because the common Python data structures are all mutable.}

\subsubsection{Unboxed Primitive}

>>>>

The \SP{} solution is to provide two versions of every parameterized type,
both of which support efficient run-time checks.
One version is shallow.
The other is \emph{concrete}: complete, but new.
Consider \code{List[T]} as an example:
\begin{enumerate}
  \item
    \emph{Shallow type \code{List}}.
    The default type \code{List[T]} gets enforced with a simple check for lists that ignores
    the elements.
    Consequently, every static type like \code{List[int]} is merely a promise of \code{List}
    values at run-time.
  \item
    \emph{Concrete type \code{ChkList[T]}}.
    \SP{} introduces a new data structure and parameterized type \code{ChkList[T]}
    for concrete types.
    The data structure implements the list API and comes with a checked constructor that
    makes it impossible to create a \code{ChkList[int]} with non-integer elements.
    Each instance of a \code{ChkList[int]} comes with the full type as a tag;
    run-time checks inspect this tag for an exact match.\footnote{Immutable
    data structures could use a subtype check. \SP{} does not have any of these
    because the common Python data structures are all mutable.}
\end{enumerate}

@@

\subsubsection{Non-Highlights, Restrictions}

The \SP{} type system does not include some features.
Each of the following is low priority, but they are rougly ordered from most
important (soonest to be implemented) to least.

@@

  \item \emph{Recursive Types}.
    \SP{} does not support recursive types.
    Neither does Pyre or mypy.
    %% demand seems low ... baffling
\end{itemize}

\subsection{Runtime System Highlights}

\begin{itemize}
  \item
    \textbf{VTables}
    Cinder adds virtual method tables to typed classes.
    ``The v-tables are only used for the \code{INVOKE\_METHOD} opcode, which
    the static compiler will only emit against a statically known static type.'' -Carl

@@

  \item
    \textbf{Bytecode Optimization}
    (could use a full section)

  %\item
  %  \textbf{Shadow Frames}
  %  (CinderDoc/static_python.rst)
  %  magic import that signals to the Static
  %  Python compiler to enable “shadow frame” mode in the Cinder JIT. This
  %  improves performance of function calls by avoiding the creation of full
  %  Python frame objects until they are definitely needed (e.g. if an
  %  exception is raised.) In the future this should become default.

  \item
    \textbf{Checked Dict}
    Implementation too, to support the types.

\subsection{Additional Features}

% things we did not model ... is that worth making a distinction here?

strict modules

decorator to force inlining, for 1-line functions

\subsubsection{Primitive Types}
